\bf Approachable case studies support learning and reproducibility in data science: An example from evolutionary biology

### 2022-02-15 jsdse/Manuscript.tex
more on barriers

Using an example within evolutionary biology, we identify aspects of research workflows
that make them unapproachable to the general audience:
use of highly specialized computing jargon and programming techniques;
high cognitive load;
unspecified, unclear or lengthy goals;
content-focused descriptions instead of user-focused;
inflexible learning environment;
unapproachable (cold or intimidating) language; and
little to no diversity of representation of information.
Then, we propose a set of principles to improve the unapproachable aspects of
research workflows, and illustrate their application in a case study from
evolutionary biology that we used as teaching material.
Finally, we elaborate on the general application of these principles for
documenting research workflows and products, to provide present learners
and future researchers with tools for successful scientific reproducibility.
%> ** add a sentence on implications for teaching reproducibility

@@

Research reproducibility --the extent to which consistent results are obtained when
a scientific experiment or research workflow is repeated \citep{repdef2021}--
is a key aspect of the advancement
of science, as it constitutes a minimum standard that allows understanding research products
(e.g., methods, data, analysis, results, etc.; \cite{piwowar2013value})
to determine their reliability and generality, and eventually build up scientific
knowledge and applications based on those products
\citep{king1995replication, peng2011reproducible, powers2019open}.
In the natural sciences, rates of reproducibility are low \citep{ioannidis2005most, prinz2011believe},
which has elicited concerns about a crisis in the field \citep{baker2016reproducibility}.

@@

@@

All of OpenTree resources
%% (the stored phylogenies, the taxonomic standard, the synthetic tree, and all
%% tools developed to process and summarize stored phylogenies),
are free of cost to any user, and are available for download and use through its Graphical User
Interface (aka, a website or application that allows users to access computer functionalities, in this case OpenTree resources, with mouse or keyboard clicks).
However, reducing as many manual steps as possible in research workflows is key
for reproducibility, as manual data manipulation scales poorly and is prone to
error \citep{bakken2019journey}.
OpenTree's resources are also programmatically available through its Application
Programming Interface services (APIs; aka computer code that implements computer functionalities,
in this case OpenTree resources, that can be used by programmers
to build more functionalities),
which provide scalability and reproducibility \citep{opentreeAPIv3}.
However, this comes at a high cost for the user, which requires considerable more
computer programming experience and literacy to be able to successfully use APIs.
%% In recent years, OpenTree API services have been wrapped into software packages for R and Python
%% \citep{michonneau2016rotl, mctavish2021opentree}, open source, free of cost
%% programming languages that represent two of the most widely used programming languages
%% in the sciences \citep{baker2017scientific}. These packages should contribute to making
%% OpenTree resources more accessible to a wider user audience.
The R and Python programming languages are open source and free of cost, and represent
two of the most widely used programming languages in the sciences today \citep{baker2017scientific}.
The rotl R package \citep{michonneau2016rotl} and the OpenTree Python module
\citep{mctavish2021opentree} are now available as wrappers for OpenTree's API services.
As such, these packages should contribute to making OpenTree resources more accessible
to a wider programming user audience.

However, while learners in the natural sciences have been engaging independently
with R and Python programming languages, computer programming is not traditionally
a core skill formally taught to biologists and naturalists
\citep{sayres2018bioinformatics, wright2019the, williams2019barriers}.
As computers continue to play a larger role in most scientific disciplines \citep{piccolo2016tools},
higher baseline computational skills are required across all natural sciences.
Thus, efforts to increase reproducibility rates in the natural sciences must consider
the specifics of scientific workflows that rely on usage of computer tools and
programming languages \citep{peng2011reproducible, sandve2013ten, powers2019open}.

In the next section que describe the barriers to approachability that we identified
on our case study. We address these barriers by developing teaching materials that are
available at
https://mctavishlab.github.io/R\_OpenTree\_tutorials/.

@@

Cite carpentries and Mosi

Good primary documentation for code is thorough. It describes general usage of individual functions,
the components and variables a function can take, and it should be accompanied with
function usage examples on how to apply it \citep{karimzadeh2018top}.
As opposed to code, primary documentation is written in natural language (i.e.,
any known human language, e.g., English, Spanish, Chinese).
Primary documentation is viewed as a key element for success of a piece of code
\citep{karimzadeh2018top},
which might be why it is also usually written using highly specialized computational jargon (i.e., computationally specific concepts,
words, and phrases) as well as formal scientific language.
While this might be important for formal acceptance of the code by the scientific
and academic community, it often slows down or even
obstructs examination, application, and adoption of code by the general audience \citep{ball2017its}.

%% Vignettes and tutorials work as secondary pieces of documentation, that help to demonstrate
%% additional cases of individual function usage, and showcase function associations
%% that work for specific analysis workflows in more detail.
%% As secondary documentation has become more common practice and is more flexible
%% in its form and content, it constitutes an ideal canvas to develop, implement and
%% test principles that can overcome current barriers to code accessibility.

Some principles of an inclusive syllabus:

- high cognitive load/ skill level; (literate programming)

- unspecified, unclear or lengthy goals;(literate programming)

- content-focused descriptions instead of user-focused; (demonstrate errors)

- inflexible learning environment; (make it stable so users can learn whenever is best for them)

- unapproachable (cold or intimidating) language, use of highly specialized computing jargon; (use friendly and informal but respectful language)

- and little to no diversity of representation of information (say the same thing with different images, metaphors, stories)

@@

\subsubsection*{a. Literate programming: Demonstrate code usage with integrative examples}

% approachable practice: reduced cognitive load and specific and clear goals

@@

A story-like narrative that links pieces of code together and invites learners
to try the code, can lead learners to remember what they are doing and
why they are doing it.
This ``literate programming'' paradigm \citep{knuth1984literate, fritzson2002mathmodelica}
makes code more approachable, as it integrates narratives with computer code in
the same document, supporting learners in actively following
the code usage, supporting memory and understanding \citep{piccolo2016tools}.

We propose that documents developed with ``literate programming'' can be made more
accessible by choosing narratives that are relatable to a more general audience.
An easy way to do this in biology is choosing as model organism a charismatic taxon that a
research group is more interested in studying. For the non-specialized user base,
a highly charismatic group such as dinosaurs will do the trick.

@@

\subsubsection*{b. Demonstrate errors and warnings thoroughly}

% approachable practice: user focused examples (instead of content focus)

@@

\subsubsection*{c. Avoid jargon and expert language}

% approachable best practice: friendly and relatable language

Besides avoiding formal language, and incorporating elements of pop culture, such as picture
character icons known as ``emojis``, to make the language more familiar to a
broader target audience (see Figure \ref{fig:error}), we made an effort to specifically
complement the primary documentation by identifying
computational concepts that were assumed or were not explained in depth.
We vetted the tutorials through feedback from workshop participants as well as individual users.
We choose examples that are charismatic for the audience.
For example, when we presented the tutorial for a team specialized in Amphibians,
we tailored the examples using frogs and their allies.

\subsubsection*{d. Make it stable through time}

>>>>

Using an example within evolutionary biology, we identify aspects of research workflows
that make them unapproachable to the general audience:
use of specialized computing language and programming techniques;
high cognitive load;
unspecified, unclear or lengthy goals;
content-focused descriptions instead of user-focused;
inflexible learning environment;
unapproachable (cold or intimidating) language; and
little to no diversity of representation of information.
Then, we propose a set of principles to improve the unapproachable aspects of
research workflows, and illustrate their application in a case study from
evolutionary biology that we used as teaching material.
Finally, we elaborate on the general application of these principles for
documenting research workflows and products, to provide present learners
and future researchers with tools for successful scientific reproducibility.
%> ** add a sentence on implications for teaching reproducibility

@@

Research reproducibility --the extent to which consistent results are obtained when
a scientific experiment or research workflow is repeated \citep{repdef2021}--
is a key aspect of the advancement
of science, as it constitutes a minimum standard that allows understanding research products,
i.e., methods, data, analysis, results, etc. \citep{piwowar2013value},
to determine their reliability and generality, and eventually build up scientific
knowledge and applications based on those products
\citep{king1995replication, peng2011reproducible, powers2019open}.
In the natural sciences, rates of reproducibility are low \citep{ioannidis2005most, prinz2011believe},
which has elicited concerns about a crisis in the field \citep{baker2016reproducibility}.

@@

\bigskip
\bigskip

@@

All of OpenTree resources
%% (the stored phylogenies, the taxonomic standard, the synthetic tree, and all
%% tools developed to process and summarize stored phylogenies),
are free of cost to any user, and are available for download and use through its Graphical User
Interface (GUI; aka, a website or application that allows users to access computer functionalities, in this case OpenTree resources, with mouse or keyboard clicks).
However, reducing as many manual steps as possible in research workflows is key
for reproducibility, as manual data manipulation scales poorly and is prone to
error \citep{bakken2019journey}.
OpenTree's resources are also programmatically available through its Application
Programming Interface services (APIs; aka, computer code that implements computer functionalities,
in this case OpenTree resources, that can be used by programmers
to build more functionalities),
which provide scalability and reproducibility \citep{opentreeAPIv3}.
However, this comes at a high cost for the user, which requires considerable more
computer programming experience and literacy to be able to successfully use APIs.
%% In recent years, OpenTree API services have been wrapped into software packages for R and Python
%% \citep{michonneau2016rotl, mctavish2021opentree}, open source, free of cost
%% programming languages that represent two of the most widely used programming languages
%% in the sciences \citep{baker2017scientific}. These packages should contribute to making
%% OpenTree resources more accessible to a wider user audience.
The \texttt{rotl} R package \citep{michonneau2016rotl} and the \texttt{opentree} Python module
\citep{mctavish2021opentree} have been developed as wrappers for OpenTree's API services.
R and Python are open source and free of cost programming languages that represent
two of the most widely used programming languages in the sciences today \citep{eglen2009quick, baker2017scientific}.
As such, \texttt{rotl} and \texttt{opentree} software packages should contribute
to making OpenTree resources more accessible to a wider programming user audience.

However, while learners in the natural sciences have been engaging independently
with R and Python programming languages, computer programming is not traditionally
a core skill formally taught to biologists and naturalists
\citep{sayres2018bioinformatics, wright2019the, williams2019barriers}.
As computers continue to play a larger role in most scientific disciplines \citep{piccolo2016tools},
higher baseline computational skills are required across all natural sciences not only
to develop an original research workflow, but to be able to follow and reproduce
research workflows from other researchers.

Thus, efforts to increase reproducibility rates long term in the natural sciences
would benefit from addressing specific barriers for learners in the field, to
support them in acquiring the skills needed to reproduce research workflows
that rely heavily on computer code \citep{peng2011reproducible, sandve2013ten, powers2019open}.

In the next section, we describe the barriers to approachable research workflows that we identified
on our case study. We address these barriers in a set of teaching materials that are
available at
https://mctavishlab.github.io/R\_OpenTree\_tutorials/.

@@

% Cite carpentries and Mosi

The main goal of our research workflow is to obtain a single phylogeny summarizing
data from a set of published phylogenies for the canids (the family of dogs, coyotes, wolves, etc.),
our organism of study.
Our analysis can be completely accomplished using functions
from the R package \texttt{rotl} or the Python module \texttt{opentree}. If
a researcher were to use this workflow in a publication, they would typically
describe it in the methodology section as ``The canid summary phylogeny was obtained using
functions from X package, details are available as supplementary files".
This is usual practice, mainly because journals do not have space to publish
all code used in an analysis in the methods section. Yet, supplementary files have
the misfortune to not be peer-reviewed as thoroughly (or at all) as the main manuscript
\citep{CITATION}.
They are also prone to the dreaded promise ``available upon request", which has
very low rates of fulfillment \citep{CITATION}.
Without the primary code that was used to perform an analysis it is impossible
to reproduce said analysis. When the code is available, other issues can
complicate reproduction of the analysis to the point of completely obstructing
reproducibility.
For example, what software do I need to open the code files? Can I run the code within the same
software, or do I need a different one?
Do I need additional software that the analysis depends on? What does the code even mean?

These questions are usually addressed in the software documentation.
As opposed to code, documentation is written in natural language (i.e.,
any known human language, e.g., English, Spanish, Chinese), and is considered
a key element for successful adoption of software
by the target users \citep{karimzadeh2018top}. This might explain
why documentation for software addressed to academic users is also
usually written using highly specialized computational
language or jargon (i.e., computationally specific concepts,
words, and phrases) as well as formal scientific and academic language.
\textit{Barrier 1.--} While scientific jargon might have an important role for formal acceptance
of software by the scientific and academic community, it can be perceived as
cold and/or intimidating language, that often slows down or even
obstructs examination, application, and adoption of code by a wider audience \citep{ball2017its},
and discourages learners by creating a hostile environment that does not foster learning \citep{CITATION}.

Good software documentation for code has to be thorough \citep{karimzadeh2018top}.
It should describe general usage of individual functions, as well as
the argument and variables said function can take, and it should be accompanied with
usage examples on how to apply each function \citep{karimzadeh2018top}.
Individual documentation for each function is usually presented
in alphabetic order, and does not have a specific goal, besides being as thorough as possible.
In this context, identifying connections across functions that are meant to work on the same
analysis workflow is difficult. Moreover, most software has many functions, so documentation
is usually very lengthy. \textit{Barrier 2.--}  All of this increases the cognitive
load for the target users and learners, which discourages learners \citep{CITATION}.

Another important aspect of software documentation is that examples are usually
worked and showcase the ideal or minimal case in which a function should work well.
\textit{Barrier 3.--} Focusing importance on the content of the software instead
of considering the user experience, ignoring what can go wrong and avoiding clear
advice on how to troubleshoot.

Software documentation is usually explained with words, very rarely using diagrams or
other allegories.
\textit{Barrier 4.--} Little to no diversity of representation of information

Documentation should always be available for the users and learners
\textit{Barrier 5.--} Inflexible learning environment, is this one really an issue???

% Some principles of an inclusive syllabus:

% - unspecified, unclear or lengthy goals; high cognitive load; (literate programming)

% - content-focused descriptions/examples instead of user-focused; (demonstrate errors)

% - inflexible learning environment; (make it stable so users can learn whenever is best for them)

% - unapproachable (cold or intimidating) language, use of highly specialized computing jargon, skill level; (use friendly and informal but respectful language)

% - and little to no diversity of representation of information (say the same thing with different images, metaphors, stories)

In sum, best practices for good primary documentation are not enough to ensure
reproducibility of research workflows that rely heavily on code.

@@

\subsubsection*{a. Reduce cognitive load and provide specific and clear goals
with literate programming}

@@

A story-like narrative that links code usage in an integrative example, invites learners
to try the code, which can lead them to remember what they are doing and
why they are doing it.
This ``literate programming'' paradigm \citep{knuth1984literate, fritzson2002mathmodelica}
makes code more approachable, as it integrates narratives with computer code in
the same document, supporting learners in actively following
the code usage, supporting memory and understanding \citep{piccolo2016tools}.

We propose that documents developed with ``literate programming'' can be made more
accessible by choosing narratives that are relatable to a more general audience.
An easy way to do this in biology is choosing a charismatic taxon as model organism.
For a research group this can be the group they are studying. For the general audience,
a highly charismatic group such as dinosaurs will do the trick.

@@

\subsubsection*{b. Provide examples that are user-focused by demonstrating errors and warnings}

% approachable practice: user focused examples (instead of content focus)
% demonstrate errors and warnings thoroughly

@@

\subsubsection*{c. Use friendly, relatable and respectful language}

% avoid specialized, intimidating and cold language

Besides avoiding formal language, and incorporating elements of pop culture, such as picture
character icons known as ``emojis``, to make the language more familiar to a
broader target audience (see Figure \ref{fig:error}), we made an effort to specifically
complement the primary documentation by identifying
computational concepts that were assumed or were not explained in depth.
We vetted the tutorials through feedback from workshop participants as well as
individual users to identify such specialized concepts.

\subsubsection*{d. Make it accessible geographically and through time}

% - inflexible learning environment; (make it stable so users can learn whenever is best for them)
